---
title: "Introduction to for loops in R"
author: "Will Cornwell, Daniel Falster"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this prac is to learn to get the computer to repeat things for you and therefore avoid tedium and burnout and stay a happy PhD student with a obedient computer rather than a frustrated, burned-out one with RSI. 
```{r}
library(tidyverse)

for (current_number in c(3, 7, 3)) {
  print(current_number)
}
5 + 3

c(3, 7, 3)[2]


letters
```

# Your first for-loop

There are datasets in R already.  one of them is called `letters`.  It does this

```{r}
letters

LETTERS
```

keep this in mind for the tasks below.

**Exercises:**
1. Write a `for` loop that prints out each letter sequentially through the alphabet

```{r}
for (alphabet in letters) {
  print(alphabet)
}
```

2. (stretch) Write a `for` loop that prints out every letter except `q`.  never liked `q` anyway.  

```{r}
for (alphabet in letters) {
    if (alphabet == 'q'){
      next
    }
        print(alphabet)
}

#OR

for (alphabet in letters) {
    if (alphabet != 'q')
        print(alphabet)
}

#OR

for (alphabet in letters[letters != "q"]) {
        print(alphabet)
}
```


Now some fun code (which will also be useful later). This code makes a plot of the world.  You can see that it's just a `ggplot` call with a `geom_polygon` added.  Then the `coord_fixed` call is just scaling x relative y so that the world is not distorted.  This is a common thing to do with maps.

```{r}
world <- map_data("world")
ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3) +
  theme_bw()
```

Note that we're putting the map data into the `geom` part of this call.  This means that we can add additional data to other geoms and plot on top of the map 

```{r}
origin <- data.frame(x = 0, y = 0)

ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group)) +
  coord_fixed(1.3) +
  geom_point(data = origin, aes(x = x, y = 0), col = "red", size = 5)
```

Now here is a new set of data on penguins! ðŸ§ðŸ§ðŸ§ðŸ§ and maybe some new functions to learn

these are real data from GBIF with all of the features and flaws of real data.  

```{r}
csv_file_name_vector <- list.files("data/penguins/", full.names = TRUE)
# note that this returns the path from the working directory to all the penguin files.
# apparently there are 19 penguin species in the world.  The first one alphabetically is Aptenodytes forsteri
```

https://en.wikipedia.org/wiki/Emperor_penguin


```{r}
a <- read_csv("data/penguins/Aptenodytes forsteri.csv")
a

# Putting all penguins in a list

list_of_data <- list()

aa <- read_csv(csv_file_name_vector[1])
list_of_data[[1]] <- read_csv(csv_file_name_vector[1])
list_of_data[[2]] <- read_csv(csv_file_name_vector[2])

dim(list_of_data[[1]])
#[1] 330  50
dim(list_of_data[[2]])
#[1] 3430   50

colnames(list_of_data[[1]]) == colnames(list_of_data[[2]]) #same columns, we can bind them

output_df <- bind_rows(list_of_data)

dim(output_df)
```

Ugh this is gonna take forever ... what about a loop?

```{r}
files <- list.files("data/penguins/", full.names = TRUE) #full.names is necessary otherwise it saves the names of csv files in penguins in a vector names files
```

sometimes you want to use the index in the `for` loop for two different things, (1) going through the input vector and (2) storing in the output list. In this case the convention is to use the letter `i`, because it's hard to think of a more informative way to describe these two different (but related) roles.  The vector that `i` goes over is *usually* from 1 to the number of inputs or in this case `1:length(files)`

```{r}
pen <- list()
for (i in 1:length(files)) {
  pen[[i]] <- read_csv(files[i])
}

class(pen) #checking what type of elemtn pen is
# [1] "list"]
pen_all <- bind_rows(pen)
dim(pen_all)
```

That's it. you're now all `for` loop experts. Time for practice.  

## More Exercises

1. Calculate how many species, genera, families, and orders there are in this dataset

```{r}
n_distinct(pen_all$species)
#[1] 19
n_distinct(pen_all$genus)
#[1] 6
n_distinct(pen_all$family)
#[1] 1
n_distinct(pen_all$order)
#[1] 1
```

2. calculate the proportion of points from iNaturalist and eBird using the institutionCode (the code for eBird is confusingly "CLO"; the code for iNaturalist is more sensibly "iNaturalist")

```{r}
clo_count <- pen_all %>% 
  filter(institutionCode == "CLO") %>% 
  summarise(count = n())
inat_count <- pen_all %>% 
  filter(institutionCode == "iNaturalist") %>% 
  summarise(count = n())
(clo_count/(inat_count+clo_count))*100
(inat_count/(inat_count+clo_count))*100

```

3. plot the geographic distribution of observations with the colors showing genus, try to make it look good.  (this blog might help https://sarahpenir.github.io/r/making-maps/)

```{r}
worldmap <- ggplot() +
  geom_polygon(data = world, aes(x=long, 
                          y = lat,
                          group = group)) + 
  coord_fixed(1.3)

genus_map <- worldmap +
  geom_point(data = pen_all, aes(x = decimalLongitude, y = decimalLatitude, col=genus), size = 5)

genus_map
```

5. write a for loop that makes a map of each genus in a separate file with species as the colors.  Save the maps in a folder called `penguin_maps`.  `ggsave` is a handy function.  

```{r}
pen_genera <- unique(pen_all$genus)

for (g in pen_genera) {
  data_genus <- pen_all %>% 
    filter(genus == g)
  
p <- ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3) +
  geom_point(data = data_genus, aes(x = decimalLongitude, y = decimalLatitude, col=species), size = 5)
    
ggsave(paste0(g, ".png"), p, path = "output")
}

```

6. plot the geographic distribution of observations of more than 1000 penguins, with the colors showing number of penguins observed

```{r}
unique(pen_all$individualCount)

pen_1000 <- unique(pen_all$individualCount)

for (h in pen_1000) {
  data_1000 <- pen_all %>% 
    filter(individualCount > 1000)
  
t <- ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3) +
  geom_point(data = data_1000, aes(x = decimalLongitude, y = decimalLatitude, col= individualCount), size = 3) +
  scale_color_continuous(limits = c(1000, max(data_1000$individualCount)))
    
#ggsave(paste0(h, ".png"), t, path = "output")
}
```


7. find a dubious penguin record and identify who's to blame
```{r}
for (variable in vector) {
  
}
```

8. Filter the data set to only iNaturalist records (see column institutionCode), split by species, and write each species data to a separate file in a folder called `inat_penguins`  You'll also need a function called `paste` or `paste0`

Note in this case the index is only needed for one thing so you can use an informative index.

Here is part of the answer: the line of code for writing the dataframe to a file, with one way to do the naming part.  

```{r}
write_csv(curr_df, paste0("data/inat_penguins/", current_species,".csv"))
```

see if you can figure out the rest.  Other functions you might need are `unique` and `filter`, but there are many different solutions

```{r}

```
